---
---

```{r, echo = FALSE}

exercise_number <- 0
exercise <- function(txt) {
  exercise_number <<- exercise_number + 1
  cat(paste0("\n**Exercise** ", exercise_number), paste0("*", txt, "*\n"))
}

```


# Introduction

> ## Learning Objectives {.objectives}
>
> * To gain familiarity with the various panes in the RStudio IDE
> * To gain familiarity with the buttons, short cuts and options in the RStudio IDE
> * To understand variables and how to assign to them
> * To be able to manage your workspace in an interactive R session
> * To be able to use mathematical and comparison operations
> * To be able to call functions
> * Introduction to package management
> 

## Introduction to RStudio

Throughout this lesson, we're going to teach you some of the fundamentals of
the R language as well as some best practices for organising code for
scientific projects that will make your life easier.

We'll be using RStudio: a free, open source R integrated development
environment. It provides a built in editor, works on all platforms (including
on servers) and provides many advantages such as integration with version
control and project management.

**Basic layout**

When you first open RStudio, you will be greeted by three panels:

  * The interactive R console (entire left)
  * Environment/History (tabbed in upper right)
  * Files/Plots/Packages/Help/Viewer (tabbed in lower right)

Once you open files, such as R scripts, an editor panel will also open
in the top left.

## Work flow within RStudio
There are two main ways one can work within RStudio.

1. Test and play within the interactive R console
   *  This works well when doing small tests and initially starting off.
   *  It quickly becomes laborious
2. Start writing in an .R file and use RStudio's command / short cut
to push current line, selected lines or modified lines to the
interactive R console. 
   * This is a great way to start; all your code is saved for later
   * You will be able to run the file you create from within RStudio
   or using R's `source()`  function.

> ## Tip: Running segments of your code {.callout}
>
> RStudio offers you great flexibility in running code from within the editor
> window. There are buttons, menu choices, and keyboard shortcuts. To run the
> current line, you can 1. click on the `Run` button just above the editor panel,
> or 2. select "Run Lines" from the "Code" menu, or 3. hit Ctrl-Enter in Windows
> or Linux or Command-Enter on OS X. (This shortcut can also be seen by hovering
> the mouse over the button). To run a block of code, select it and then `Run`.
> If you have modified a line of code within a block of code you have just run,
> there is no need to reselct the section and `Run`, you can use the next button
> along, `Re-run the previous region`. This will run the previous code block
> inculding the modifications you have made.

## Introduction to R

Much of your time in R will be spent in the R interactive
console. This is where you will run all of your code, and can be a
useful environment to try out ideas before adding them to an R script
file. This console in RStudio is the same as the one when you open up the basic R GUI.

The first thing you will see in the R interactive session is a bunch
of information, followed by a ">" and a blinking cursor. It operates on the idea 
of a "Read, Evaluate, Print loop" (REPL): you type in commands, R tries to execute them, and then
returns a result.

## Using R as a calculator

The simplest thing you could do with R is do arithmetic:

```{r}
1 + 100
```

And R will print out the answer, with a preceding "[1]". Don't worry about this
for now, we'll explain that later. For now think of it as indicating ouput.

If you type in an incomplete command, R will wait for you to
complete it:

~~~ {.r}
> 1 +
~~~

~~~ {.output}
+
~~~

Any time you hit return and the R session shows a "+" instead of a ">", it
means it's waiting for you to complete the command. If you want to cancel
a command you can simply hit "Esc" and RStudio will give you back the ">"
prompt.

> ## Tip: Cancelling commands {.callout}
>
> Cancelling a command isn't just useful for killing incomplete commands:
> you can also use it to tell R to stop running code (for example if its
> taking much longer than you expect), or to get rid of the code you're
> currently writing.
>

When using R as a calculator, the order of operations is the same as you
would have learnt back in school.

From highest to lowest precedence:

 * Parentheses: `(`, `)`
 * Exponents: `^` or `*`*
 * Divide: `/`
 * Multiply: `*`
 * Add: `+`
 * Subtract: `-`

```{r}
3 + 5 * 2
```

Use parentheses to group operations in order to force the order of
evaluation if it differs from the default, or to make clear what you 
intend.

```{r}
(3 + 5) * 2
```

This can get unwieldy when not needed, but  clarifies your intentions.
Remember that others may later read your code.

```{r, eval=FALSE}
(3 + (5 * (2 ^ 2))) # hard to read
3 + 5 * 2 ^ 2       # clear, if you remember the rules
3 + 5 * (2 ^ 2)     # if you forget some rules, this might help
```


The text after each line of code is called a
"comment". Anything that follows after the hash (or octothorpe) symbol
`#` is ignored by R when it executes code.

Really small or large numbers get a scientific notation:

```{r}
2 / 10000
```

Which is shorthand for "multiplied by `10^XX`". So `2e-4`
is shorthand for `2 * 10^(-4)`.

You can write numbers in scientific notation too:

```{r}
5e3  # Note the lack of minus here
```

## Functions
Most of R's functionality comes from its *functions*. A function takes zero, one or multiple *arguments*, depending on the function, and returns a value. To call a function enter it's name followed by a pair of brackets - include any arguments in the brackets.
```{r}
log(10)
```

To find out more about a function called `function_name` type `?function_name`. To search for the functions associated with a topic type `??topic` or `??"multiple topics"`. As well as providing a detailed description of
the command and how it works, scrolling ot the bottom of the
help page will usually show a collection of code examples which
illustrate command usage. 

```{r, echo = FALSE, results = 'asis'}
exercise("Which function calculates sums? And what arguments does it take?")
```

## Arguments
The documentation for `log` indicates that the function requires an argument `x` that is a vector of
*numeric* (real) or *complex* numbers and an argument `base` which is the base of the logarithm. 

```{r, echo = FALSE, results = 'asis'}
exercise("What kind of logarithm does the `log` function take by default?")
```

When calling a function its arguments can be specified using *positional* and/or 
*named* matching.

```{r}
log(x = 10, base = 2)
log(10, 2)
log(2, 10)
```

## Mathematical functions

R has many built in mathematical functions.

```{r}
sin(1)  # trigonometry functions
```

```{r}
log(1)  # natural logarithm
```

```{r}
log10(10) # base-10 logarithm
```

```{r}
exp(0.5) # e^(1/2)
```

Don't worry about trying to remember every function in R. You
can simply look them up on google, or if you can remember the
start of the function's name, use the tab completion in RStudio.

This is one advantage that RStudio has over R on its own, it
has autocompletion abilities that allow you to more easily
look up functions, their arguments, and the values that they
take.

## Comparing things

We can also do comparison in R:

```{r}
1 == 1  # equality (note two equals signs, read as "is equal to")
```

```{r}
1 != 2  # inequality (read as "is not equal to")
```

```{r}
1 <  2  # less than
```

```{r}
1 <= 1  # less than or equal to
```

```{r}
1 > 0  # greater than
```

```{r}
1 >= -9 # greater than or equal to
```

> ## Tip: Comparing Numbers {.callout}
>
> A word of warning about comparing numbers: you should
> never use `==` to compare two numbers unless they are
> integers (a data type which can specifically represent
> only whole numbers).
>
> Computers may only represent decimal numbers with a
> certain degree of precision, so two numbers which look
> the same when printed out by R, may actually have
> different underlying representations and therefore be
> different by a small margin of error (called Machine
> numeric tolerance).
>
> Instead you should use the `all.equal` function.
>
> Further reading: [http://floating-point-gui.de/](http://floating-point-gui.de/)
>

## Variables and assignment

We can store values in variables by giving them a name, and using the assignment operator `<-` (To save finger strokes, type `Alt-`):

```{r}
x <- 1 / 40
```

Notice that assignment does not print a value. Instead, we stored it for later
in something called a **variable**. `x` now contains the **value** `0.025`:

```{r}
x
```

Look for the `Environment` tab in one of the panes of RStudio, and you will see that `x` and its value
have appeared. Our variable `x` can be used in place of a number in any calculation that expects a number:

```{r}
log(x)
```

Notice also that variables can be reassigned:

```{r}
x <- 100
```

`x` used to contain the value 0.025 and and now it has the value 100.

Assignment values can contain the variable being assigned to:

```{r}
x <- x + 1 #notice how RStudio updates its description of x on the top right tab
```

The right hand side of the assignment can be any valid R expression.
The right hand side is *fully evaluated* before the assignment occurs.

```{r, echo = FALSE, results = 'asis'}
exercise("Create an object called `x` with the value `7`. What is the value of `x^x`. Save the value in a object called `i`. If you assign the value `20` to the object `x` does the value of `i` change? What does this indicate about how R assigns values to objects?")
```

Variable names can contain letters, numbers, underscores and periods. They
cannot start with a number nor contain spaces at all. Different people use
different conventions for long variable names, these include

  * periods.between.words
  * underscores\_between_words
  * camelCaseToSeparateWords

What you use is up to you, but **be consistent**.

It is also possible to use the `=` operator for assignment:

```{r}
x = 1 / 40
```

But this is much less common among R users.  The most important thing is to
**be consistent** with the operator you use. There are occasionally places
where it is less confusing to use `<-` than `=`, and it is the most common
symbol used in the community. So the recommendation is to use `<-`.

## Vectorization

One final thing to be aware of is that R is *vectorized*, meaning that
variables and functions can have vectors as values. For example

```{r}
1:5
(1:5)^2
x <- 1:5
x^2
```

This is incredibly powerful; we will discuss this further in an
upcoming lesson.


## Managing your environment

There are a few useful commands you can use to interact with the R session.

`ls` will list all of the variables and functions stored in the global environment
(your working R session):

```{r}
ls()
```

Note here that we didn't given any arguments to `ls`, but we still
needed to give the parentheses to tell R to call the function.

If we type `ls` by itself, R will print out the source code for that function!

```{r}
ls
```

You can use `rm` to delete objects you no longer need:

```{r}
rm(x)
```

If you have lots of things in your environment and want to delete all of them,
you can pass the results of `ls` to the `rm` function:

```{r}
rm(list = ls())
```

In this case we've combined the two. Just like the order of operations, anything
inside the innermost parentheses is evaluated first, and so on.

In this case we've specified that the results of `ls` should be used for the
`list` argument in `rm`.

> ## Tip: Warnings vs. Errors {.callout}
>
> Pay attention when R does something unexpected! Errors, like above,
> are thrown when R cannot proceed with a calculation. Warnings on the
> other hand usually mean that the function has run, but it probably
> hasn't worked as expected.
>
> In both cases, the message that R prints out usually give you clues
> how to fix a problem.
>

> ## Challenge 1 {.challenge}
> Which of the following are valid R variable names?
> ```{r, eval=FALSE}
> min_height
> max.height
> _age
> .mass
> MaxLength
> min-length
> 2widths
> celsius2kelvin
> ```

> ## Challenge 2 {.challenge}
>
> What will be the value of each  variable  after each
> statement in the following program?
>
> ```{r, eval=FALSE}
> mass <- 47.5
> age <- 122
> mass <- mass * 2.3
> age <- age - 20
> ```
>

> ## Challenge 3 {.challenge}
>
> Run the code from the previous challenge, and write a command to
> compare mass to age. Is mass larger than age?
>

> ## Challenge 4 {.challenge}
>
> Clean up your working environment by deleting the mass and age
> variables.
>

** The structure and much of the content in this module was borrowed from: 
- [Software Carpentry's](http://www.swcarpentry.org)'s [R Gapminder lesson](http://swcarpentry.github.io/r-novice-gapminder/index.html), and
- [Poisson Consulting](http://www.poissonconsulting.ca)'s
[Introductory course notes](http://www.poissonconsulting.ca/course/2014/09/12/an-introduction-to-r-course.html), 
- Both are available for reuse under the Creative Commons Attribution 
[(CC_BY)](http://creativecommons.org/licenses/by/3.0/) 
license.

