---
title: Getting started with data
---

```{r, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(results="markup")
```

## Analyzing Iris Data

We are studying differences in morphology in two species of Iris: _Iris setosa_ 
and _Iris versicolor_.

### We want to:

* load that data into memory,
* Explore the data, 
* Calculate some summary statistics, 
* Explore the dataset visually, 
* plot the result.

To do all that, we'll have to learn a little bit about programming.

### Objectives
* Read tabular data from a file into **R**.
* Assign values to variables.
* Learn about data types and structures
* Select individual values and subsections from data
* Perform operations on data frames
* Explain what a package is, and what packages are used for.
* Load an R package and use the things it contains

## RStudio Projects
* Projects in RStudio help us keep our code, data, and outputs well organized
* when we quit RStudio, it will help us remember where we left off
* Open a new project -> New Directory -> Empty Project
* Look around  - you'll see where we are
* Create a directory to hold the data

```{r}
getwd()

```

## Loading Data

To load our iris data, we need to download our data to a place we can use it
We will use `dir.create()` and `read.csv()`. These are built-in functions in R. 
Let's check out the help screen.

```{r}
dir.create("data")

```

* download the iris file
* put it in your data directory for these exercises

Load the data into R:

```{r}
read.csv('data/iris.csv')
```

The expression `read.csv()` is a function that asks R to load a `.csv` file 
into R's workspace as a data frame.

`read.csv()` has many arguments including the name of the file we want to read, 
and the delimiter that separates values on a line. 

When we are finished typing and press `Control+Enter` on Windows or 
`Cmd + Return` on Mac, the console runs our command. 
Since we haven't told it to do anything else with the function's output, the 
console displays it.
In this case, that output is the data we just loaded. 

Our call to `read.csv()` read the file, but didn't save the data as an object. 
To do that, we need to assign the data frame to a variable. 
A variable is just a name for a value, such as `x`, `current_temperature`, or 
`subject_id`. 
We can create a new variable simply by assigning a value to it using `<-`

```{r}
weight_kg <- 55
```

Once a variable has a value, we can print it:

```{r}
weight_kg
```

and do arithmetic with it:

```{r}
2.2 * weight_kg
```

We can use the function `paste()` to strings made up from different objects, if 
we separate them with commas.

```{r}
paste("weight in pounds:", 2.2 * weight_kg)
```

We can also change an object's value by assigning it a new value:

```{r}
weight_kg <- 57.5
# weight in kilograms is now
weight_kg
```

If we imagine the variable as a sticky note with a name written on it, 
assignment is like putting the sticky note on a particular value
This means that assigning a value to one object does not change the values of 
other variables.

For example, let's store the subject's weight in pounds in a variable

```{r}
weight_lb <- 2.2 * weight_kg
# weight in kg...
weight_kg
# ...and in pounds
weight_lb
```

and then change `weight_kg`:

```{r}
weight_kg <- 100.0
# weight in kg now...
weight_kg
# ...and in weight pounds still
weight_lb
```

__Updating a Variable__

Since `weight_lb` doesn't "remember" where its value came from, it isn't 
automatically updated when `weight_kg` changes. This is different from the way 
spreadsheets work.

__Challenges__

Draw diagrams showing what variables refer to what values after each statement in the following program:

```{r}	
mass <- 47.5
age  <- 122
mass <- mass * 2.0
age  <- age - 20
```

### Vectors

So far all of our objects have been single numbers... but they are acually 
vectors of length1. A vector is the most common and basic data structure in `R` 
and is pretty much the workhorse of R. It is a one-dimensional object with a 
series of values, all of the same type.

You can create, or add to vectors using the function `c()`. `c()` takes any 
number of vectors as arguments, and combines or *concatenates* them, in the 
order supplied, into a single vector. For example, if we are creating a vector 
of patient weights, we could create and then update that vector using `c()`.

```{r}
weights <- c(100, 70, 45, 62, 81)
weights <- c(weights, 80)
weights
```

What happens here is that we take the original vector weights, and we are adding 
the second item to the end of the first one. We can do this over and over again 
to build a vector or a dataset. As we program, this may be useful to autoupdate 
results that we are collecting or calculating.

`weights` is a numeric vector, but there are other kinds as well: `character` 
(letters and words) and `logical` (`TRUE`or `FALSE`) are the other most common.

#### A note about naming things:

> There are only two hard things in Computer Science: cache invalidation and naming things.
> 
> -- Phil Karlton

Name things so that you can understand them later, try to balance brevity with 
clarity. `x` is easy to type, but may not mean much.

* Only begin names with letters.
* Separate words with a dot (`my.var`) or an underscore (`my_var`), or use 
`camelCase`.
* Try to be consistent.
* Avoid giving things names that already exist (like `mean`, `sum`, `log`)

Now that we know how to assign things to variables, let's re-run `read.csv` and 
save its result:

```{r}
dat <- read.csv('data/iris.csv')
```

This statement doesn't produce any output because assignment doesn't display 
anything. If we want to check that our data has been loaded, we can print the 
variable's value:

```{r}
dat
```

It's important to note that anything we do with `dat` here will not affect the 
original `csv` file. This is an important concept of reproducibility - don't 
touch your original data!

For large data sets it is convenient to use the `head()` to display the first 
few rows of data, or `tail()` to see the end

```{r}
head(dat)
tail(dat)
```

In RStudio's environment pane, you can also click on the little table icon next
to the data frame name.

__BREAK__

* Make sure everyone has imported the data
* How many rows and columns there are
* What kind of data type is it?

## Manipulating Data

Now that our data is in memory, we can start doing things with it. 
First, let's ask what type of thing `dat` *is*:

```{r}
class(dat)
str(dat)
summary(dat)
```

The output tells us that data currently is a data frame in R. 
This is similar to a spreadsheet in MS Excel, that many of us are familiar with 
using.

### data frames
The *de facto* data structure for most tabular data and what we use for statistics.

Data frames can have additional attributes such as `rownames()`, which can be 
useful for annotating data, like `subject_id` or `sample_id`. But most of the 
time they are not used.

Some additional information on data frames:

* Usually created by `read.csv()` and `read.table()`.
* Can convert to matrix with `data.matrix()` (preferred) or `as.matrix()`
* Coercion will be forced and not always what you expect.
* Can also create with `data.frame()` function.
* Find the number of rows and columns with `nrow(dat)` and `ncol(dat)`, 
respectively.

#### Useful data frame functions

* `head()` - shown first 6 rows
* `tail()` - show last 6 rows
* `dim()` - returns the dimensions
* `nrow()` - number of rows
* `ncol()` - number of columns
* `str()` - structure of each column
* `names()` - shows the `names` attribute for a data frame, which gives the column names.

`str` output tells us the dimensions and the data types (int is integer) of each column.

We can see what its shape is like this:

```{r}
dim(dat)
nrow(dat)
ncol(dat)
```

This tells us that `dat` has `r nrow(dat)` rows and `r ncol(dat)` columns.

### Referencing Columns

When referencing a column, the `$` operator is used to separate the data frame 
and the column name. In a data frame, each column is a _vector_ of the same 
length. Let's look at the petal widths:

```{r}
dat$Petal.Width
```

Now say we want to get the mean of the petal widths, or the maximum of the sepal 
widths:

```{r}
mean(dat$Petal.Width)
max(dat$Sepal.Width)
```

Wait, what happened there? Why is the max of sepal width `NA`?  Let's have a look:

```{r}
dat$Sepal.Width # First look at the data - there are some values that are NA.
```

`NA` is a special encoding for a missing value in a vector... look at help file 
for `max()`: `?max`.  You'll see there's an argument called `na.rm`, where the 
default is `FALSE`.  If we change it to true, `max` ignores the `NA` values:

```{r}
max(dat$Sepal.Width, na.rm=TRUE)
```

You can also use the `$` operator to reference columns that don't yet exist, but
you would like to create.  For example, the measurements in this dataset are in 
cm, but perhaps we want to compare sepal lengths with another dataset that has 
measurements in inches:

```{r}
dat$Sepal.Length.in <- dat$Sepal.Length / 2.54
head(dat)
```

Or, perhaps what we are really interested in is the ratio of petal length to 
petal width:

```{r}
dat$Petal.LW.Ratio <- dat$Petal.Length / dat$Petal.Width
head(dat)
```

You can remove columns by assigning `NULL` to them. Let's get rid of those two 
we just created to get back to our original data frame.

```{r}
dat$Petal.LW.Ratio <- NULL
dat$Sepal.Length.in <- NULL
```

<!-- This may be a good place to talk about vectors -->

### Indexing

We talked about vectors before, and indexing them.  Each column in a data frame 
is a vector, and can be indexed as such.

If we want to get a single value from the data frame, we must provide an row and 
column indices for the value we want in square brackets: `dataframe[rows,columns]`

```{r}
# first value in first column in dat
dat[1, 1]
# third value in fifth column in dat
dat[3, 5]
```

An index like `dat[30, 3]` selects a single element of data frame, but we can 
select sections as well. Entering nothing in either the rows or columns spot 
selects all.
For example, we can select the first ten rows across all columns, like this. :

```{r}
dat[1:10,]
```

The notation `1:10` means, "Start at index 1 and go to index 10." 

We don't have to take consecutive values, we can use `c()` to select certain 
values or groups of values:

```{r}
dat[c(1:10, 20:30),]
```

Here we have taken rows 1 through 10 and 20 through 30.

Columns can also be referenced by name (in quotes):

```{r}
dat[, "Petal.Length"] # Returns all rows
dat[, c("Petal.Length", "Species")]  # Can specify more than one column by using a character vector

```

#### Indexing using conditionals

Sometimes we are interested in extracting data based on something else.  Lets 
say we want to look at all of the flowers measured that had really long sepals:

```{r}
mean.sl <- mean(dat$Sepal.Length) # First find the average sepal length
mean.sl
dat[dat$Sepal.Length > mean.sl, ]
```

Let's break that down a little. `dat$Sepal.Length > mean.sl` returns a _Logical_ 
vector of the same length as `dat$Sepal.Length`; that is a vector of `TRUE` and 
`FALSE` based on testing each element of `dat$Sepal.Length` to see if it is 
greater than `mean.sl`.

```{r}
dat$Sepal.Length > mean.sl
```

When you put that logical vector inside the row index, just the rows which are 
`TRUE` are returned. This type of conditional indexing is very powerful:

```{r}
dat[dat$Species == "setosa",]
```

what if you want to find the mean of the petal lengths of just _Iris setosa_?

```{r}
mean(dat[dat$Species == "setosa", "Petal.Length"])

```

## Key Points

* Import a package into your workspace using `library("pkgname")`.
* The key data types in R?
* Use `variable <- value` to assign a value to a variable in order to record it 
  in memory.
* Objects are created on demand whenever a value is assigned to them.
* Use `print(something)` (or just `something`) to display `something`.
* The expression `dim()` gives the dimensions of a data frame or matrix.
* Use `object[x, y]` to select a single element from an array.
* Object indices start at 1.
* Use `from:to` to specify a sequence that includes the indices from `from` to `to`.
* Use `#` to add comments to programs.
* Use `mean()`, `max()`, `min()` and `sd()` to calculate simple statistics.
* Create and update vectors using `c()`