---
title: Getting started with data
---

** The structure and much of the content in this module was borrowed from 
[Software Carpentry](http://www.softwarecarpentry.org)'s
[novice R Bootcamp material](https://github.com/swcarpentry/bc/tree/master/novice/r) 
(Copyright (c) Software Carpentry), which they make available for reuse under the 
Creative Commons Attribution [(CC_BY)](http://creativecommons.org/licenses/by/3.0/) 
license.

## Analyzing Iris Data

We are studying differences in morphology in two species of Iris: _Iris setosa_ 
and _Iris versicolor_.

### We want to:

* load that data into R,
* Explore the data, 
* Calculate some summary statistics, 
* Explore the dataset visually, 
* plot the results

To do all that, we'll have to learn a little bit about programming in R.

### Objectives

* Read tabular data from a spreadsheet file into **R**.
* Assign values to variables.
* Learn about data types and structures
* Select individual values and subsections from data
* Perform operations on data frames
* Explain what a package is, and what packages are used for.
* Load an R package and use the things it contains

### Working Directory

Each R session is associated with a folder on your computer that is referred to as the working directory. To view the contents of the working directory go to the Files window in the bottom right pane. To change the working directory use the 
```{r, echo = FALSE, results = 'asis'}
exercise("Create a new folder on your home drive called RCourse and make this folder the working directory. To ensure you have been successful confirm that the output of `getwd()` refers to your RCourse folder.")
```

### RStudio Projects

Instead of setting the working directory each time you restart RStudio you can associate
the contents of a folder with a *project* - then all you have to do is to open that project. 
Projects in RStudio help us keep our code, data, and outputs well organized

To create a new project use the **New Project** command (available on the **File** menu of the global toolbar).
```{r, echo = FALSE, results = 'asis'}
exercise("Create a new project called RCourse in the RCourse folder on your desktop. As the folder already exists choose the **Create project from: Existing Directory** option.
  To confirm you were successful quit RStudio and double-click the RCourse.Rproj file
  in the RCourse folder - RStudio should restart with RCourse as the working directory.")
```

## Scripts

Rather than entering expressions into R one by one in the console, 
a more efficient method is to type the expressions into a text file 
called a *script* and then send all (or a subset) of them to R at the same time. 
R scripts are indicated by the suffix `.R`.

Open a new script using the "R Script"" option of the New option in the File menu.

Now save the script in your working directory as `iris.R`.

* Lines beginning with `#` are ignored by R. Comment your code liberally using `#`.  Your future self will thank you.
* Start by typing the following at the top of your script:

```{r, eval=FALSE}
# This is my first script.  We are going to learn about irises.
```

## Loading Data

To load our iris data, we need to download our data to a place we can use it. 
We will use the function `dir.create()` to create a `data` directory in our project directory, 
and `read.csv()` to load the data file. These are built-in functions in R. Let's 
check out the help screen.

```{r, message=FALSE, warning=FALSE}
?dir.create
dir.create("data")
```

Next download the iris file to your data directory for these exercises

```{r, message=FALSE, warning=FALSE}
download.file("http://ateucher.github.io/rcourse_site/data/iris.csv", destfile = "data/iris.csv")
```

### Comma Separated Files

The simplest way to input and output data is in the form of comma separated files. Comma separated files, which have the suffix **.csv**, are recognised by almost all
statistical and spreadsheet programs including R and Excel.

Load the data into R:

```{r}
read.csv('data/iris.csv')
```

The expression `read.csv()` is a function that asks R to load the contents of a `.csv` file 
into R's workspace.

`read.csv()` has many arguments including the name of the file we want to read, 
and the delimiter that separates values on a line. 

When we are finished typing and press `Control+Enter` on Windows or 
`Cmd + Return` on Mac, the console runs our command. 
Since we haven't told it to do anything else with the function's output, the 
console displays it.
In this case, that output is the data we just loaded. 

Our call to `read.csv()` read the file, but didn't save the data as an object. 
To do that, we need to assign the data frame to a variable.

Now that we know how to assign things to variables, let's re-run `read.csv` and 
save its result:

```{r}
dat <- read.csv('data/iris.csv')
```

This statement doesn't produce any output because assignment doesn't display 
anything. If we want to check that our data has been loaded, we can print the 
variable's value:

```{r}
dat
```

It's important to note that anything we do with `dat` here will not affect the 
original `csv` file. `dat` is a copy of your data inside R. This is an important 
concept of reproducibility - don't touch your original data!

For large data sets it is convenient to use the `head()` function to display the first 
few rows of data, or `tail()` to see the end:

```{r}
head(dat)
tail(dat)
```

In RStudio's environment pane, you can also click on the little table icon next
to the data frame name.

__BREAK__

* Make sure everyone has imported the data
* How many rows and columns there are
* What kind of data type is it?

## Manipulating Data

Now that our data is in memory, we can start doing things with it. 
First, let's ask what type of thing `dat` *is*. We can do this using the `class()` 
function, which tells us what kind of *object* something is:

```{r}
class(dat)
```

The output tells us that `dat` is a data frame in R. 
This is similar to a spreadsheet in MS Excel, that many of us are familiar with 
using.

### data frames
The *de facto* data structure for most tabular data and what we use for statistics.

Some additional information on data frames:

* Usually created by `read.csv()` and `read.table()`.
* Can also create with `data.frame()` function.

Let's find out a little bit more about `dat`:

#### Useful data frame functions

* `head()` - shown first 6 rows
* `tail()` - show last 6 rows
* `dim()` - returns the dimensions
* `nrow()` - number of rows
* `ncol()` - number of columns
* `str()` - structure of of data frame - number of rows/columns and data type of each column
* `summary()` - summarises each column
* `names()` - gives the column names


```{r}
str(dat)
summary(dat)
```

We can see what its shape is like this:

```{r}
nrow(dat)
ncol(dat)
```

This tells us that `dat` has `r nrow(dat)` rows and `r ncol(dat)` columns.

### Referencing Columns

When referencing a column, the `$` operator is used to separate the data frame 
and the column name. Remember how we talked about vectors and how important they are? 
In a data frame, each column is a _vector_ of the same length. Let's look at the petal widths:

```{r}
dat$Petal.Width
```

Now say we want to get the mean of the petal widths:

```{r}
mean(dat$Petal.Width)
```

```{r, echo=FALSE, results='asis'}
exercise("Find the maximum of the sepal widths. Did anything unexpected happen?")
```

Let's have a look at the `Sepal.Width` column:

```{r}
dat$Sepal.Width # First look at the data - there are some values that are NA.
```

### Missing Values

`NA` is a special encoding for a missing value in a vector... look at help file 
for `max()`: `?max`.  You'll see there's an argument called `na.rm`, where the 
default is `FALSE`.  If we change it to true, `max` ignores the `NA` values:

```{r}
max(dat$Sepal.Width, na.rm=TRUE)
```

You can also use the `$` operator to reference columns that don't yet exist, but
you would like to create.  For example, the measurements in this dataset are in 
cm, but perhaps we want to compare sepal lengths with another dataset that has 
measurements in inches:

```{r}
dat$Sepal.Length.in <- dat$Sepal.Length / 2.54
head(dat)
```

```{r, echo=FALSE, results='asis'}
exercise("Create a new column in `dat` that contains the ratio of petal length to petal width. Call the column `Petal.Ratio`")
```

You can remove columns by assigning `NULL` to them. `NULL` is a special term that means nothing. 
Let's get rid of those two we just created to get back to our original data frame.

```{r}
dat$Petal.Ratio <- NULL
dat$Sepal.Length.in <- NULL
```

### Indexing

We talked about vectors before, and indexing them. Each column in a data frame 
is a vector, and can be indexed as such.

```{r}
dat$Petal.Width[10]
```

If we want to get a single value from the data frame, we must provide row and 
column indices for the value we want in square brackets: `dataframe[rows,columns]`

```{r}
# first value in first column in dat
dat[1, 1]
# third value in fifth column in dat
dat[3, 5]
```

An index like `dat[30, 3]` selects a single element of data frame, but we can 
select sections as well. Entering nothing in either the rows or columns spot 
selects all.
For example, we can select the first ten rows across all columns, like this. :

```{r}
dat[1:10,]
```

The notation `1:10` means, "Start at index 1 and go to index 10." 

We don't have to take consecutive values, we can use `c()` to select certain 
values or groups of values:

```{r}
dat[c(1:10, 20:30),]
```

Here we have taken rows 1 through 10 and 20 through 30.

Columns can also be referenced by name (in quotes):

```{r}
dat[, "Petal.Length"] # Returns all rows
dat[, c("Petal.Length", "Species")]  # Can specify more than one column by using a character vector

```

```{r, echo=FALSE, results='asis'}
exercise("Given what we learned about how to access a single column, and indexing vectors, what is the 19th value in the Sepal Length column? How did you find it?")
```


#### Indexing using conditionals

Sometimes we are interested in extracting data based on something else.  Lets 
say we want to look at all of the flowers measured that had really long sepals. If we look at `iris$Sepal.Length`, we can see that they tend to range from around 4-7.  So let's select the rows where `Sepal.Length` is greater than 5:

```{r}
dat[dat$Sepal.Length > 5,]
```

But let's say we want to be more sophisticated than that, and find flowers that have above average Sepal Length.

```{r}
mean.sl <- mean(dat$Sepal.Length) # First find the average sepal length
mean.sl
dat[dat$Sepal.Length > mean.sl, ]
```

Let's break that down a little. `dat$Sepal.Length > mean.sl` returns a _Logical_ 
vector of the same length as `dat$Sepal.Length`; that is a vector of `TRUE` and 
`FALSE` based on testing each element of `dat$Sepal.Length` to see if it is 
greater than `mean.sl`.

```{r}
dat$Sepal.Length > mean.sl
```

When you put that logical vector inside the row index, just the rows which are 
`TRUE` are returned. This type of conditional indexing is very powerful:

```{r}
dat[dat$Species == "setosa",]
```

what if you want to find the mean of the petal lengths of just _Iris setosa_?

```{r}
mean(dat[dat$Species == "setosa", "Petal.Length"]) #or
mean(dat$Petal.Length[dat$Species == "setosa"])
```

## Key Points

* Use `variable <- value` to assign a value to a variable in order to record it 
  in memory.
* Objects are created on demand whenever a value is assigned to them.
* Use `read.csv()` to read in a comma-delimited file (likely from a spreadsheet) to a data frame.
* Use `object[x, y]` to select a single element from a data frame.
* Use `from:to` to specify a sequence that includes the indices from `from` to `to`.
* Use `#` to add comments to programs.
* Use `mean()`, `max()`, `min()` and `sd()` to calculate simple statistics.
* Create and update vectors using `c()`