---
title: Data visualization with ggplot2
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.path = "fig/03-plotting-")
# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("data/gapminder-FiveYearData.csv", header=TRUE)
```

> ## Learning Objectives {.objectives}
>
> * To be able to use ggplot2 to generate publication quality graphics
> * To understand the basics of the grammar of graphics:
>   - The aesthetics layer
>   - The geometry layer
>   - Adding statistics
>   - Transforming scales
>   - Coloring or paneling by groups.
>

Plotting our data is one of the best ways to
quickly explore it and the various relationships
between variables.

There are three main plotting systems in R,
the [base plotting system][base], the [lattice][lattice]
package, and the [ggplot2][ggplot2] package.

[base]: http://www.statmethods.net/graphs/
[lattice]: http://www.statmethods.net/advgraphs/trellis.html
[ggplot2]: http://www.statmethods.net/advgraphs/ggplot2.html

## Base plotting

R's base (built-in) plotting functions are powerful and very flexible, but not 
overly user friendly. For simple exploratory plots that don't need to look nice, 
they are useful. They are generally specified  as `plot(x, y, ...)`

```{r baseplot1}
plot(gapminder$lifeExp, gapminder$gdpPercap)
```

You can also specify them in a formula format `plot(y ~ x, data='', ...)`
```{r baseplot2}
plot(gdpPercap ~ lifeExp, data=gapminder)
hist(gapminder$gdpPercap)
boxplot(gdpPercap ~ continent, data=gapminder)
```

## ggplot2

Today we'll be learning about the ggplot2 package developed by Hadley Wickham, because
it is the most effective for creating publication quality
graphics.

### ggplot2 and the *Grammar of Graphics*

The **ggplot2** package provides an R implementation of Leland Wilkinson's 
*Grammar of Graphics* (1999). The *Grammar of Graphics* challenges data analysts 
to think beyond the garden variety plot types (e.g. scatter-plot, barplot) and 
to consider the components that make up a plot or graphic, such as how data are 
represented on the plot (as lines, points, etc.), how variables are mapped to 
coordinates or plotting shape or colour, what transformation or statistical 
summary is required, and so on. Specifically, **ggplot2** allows users to build 
a plot layer-by-layer by specifying:

* The _data_,
* some _aesthetics_, that map variables in the data to a visual representation 
  on the plot. This tells ggplot2 how to show each variable, such as axes on the 
  plot or to size, shape, color, etc.
* a _geom_, which specifies the geometry of how the data are represented on the 
  plot (points, lines, bars, etc.),
* a _stat_, a statistical transformation or summary of the data applied prior to 
  plotting,
* _facets_, that allow the data to be divided 
  into chunks on the basis of other categorical or continuous variables and the 
  same plot drawn for each chunk.

Because **ggplot2** implements a *layered* grammar of graphics, data points and 
additional information (scatterplot smoothers, confidence bands, etc.) can be 
added to the plot via additional layers, each of which utilize further geoms, 
aesthetics, and stats.

Let's start off with an example:

```{r lifeExp-vs-gdpPercap-scatter, message=FALSE}
library(ggplot2)
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

So the first thing we do is call the `ggplot` function. This function lets R
know that we're creating a new plot, and any of the arguments we give the
`ggplot` function are the *global* options for the plot: they apply to all
layers on the plot.

We've passed in two arguments to `ggplot`. First, we tell `ggplot` what data we
want to show on our figure, in this example the gapminder data we read in
earlier. For the second argument we passed in the `aes` function, which
tells `ggplot` how variables in the **data** map to *aesthetic* properties of
the figure, in this case the **x** and **y** locations. Here we told `ggplot` we
want to plot the "lifeExp" column of the gapminder data frame on the x-axis, and
the "gdpPercap" column on the y-axis. Notice that we didn't need to explicitly
pass `aes` these columns (e.g. `x = gapminder[, "lifeExp"]`), this is because
`ggplot` is smart enough to know to look in the **data** for that column!

By itself, the call to `ggplot` isn't enough to draw a figure:

```{r}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap))
```

We need to tell `ggplot` how we want to visually represent the data, which we
do by adding a new **geom** layer. In our example, we used `geom_point`, which
tells `ggplot` we want to visually represent the relationship between **x** and
**y** as a scatterplot of points:

```{r lifeExp-vs-gdpPercap-scatter2}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

> ## Challenge 1 {.challenge}
>
> Modify the example so that the figure visualise how life expectancy has
> changed over time:
>
> ```{r, eval=FALSE}
> ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) + geom_point()
> ```
>
> Hint: the gapminder dataset has a column called "year", which should appear
> on the x-axis.
>

> ## Challenge 2 {.challenge}
>
> In the previous examples and challenge we've used the `aes` function to tell
> the scatterplot **geom** about the **x** and **y** locations of each point.
> Another *aesthetic* property we can modify is the point *color*. Modify the
> code from the previous challenge to **color** the points by the "continent"
> column. What trends do you see in the data? Are they what you expected?
>

## Geom Layers

Using a scatterplot probably isn't the best for visualising change over time.
Instead, let's tell `ggplot` to visualise the data as a line plot:

```{r lifeExp-line}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()
```

Instead of adding a `geom_point` layer, we've added a `geom_line` layer. We've
added the **by** *aesthetic*, which tells `ggplot` to draw a line for each
country.

But what if we want to visualise both lines and points on the plot? We can
simply add another layer to the plot:

```{r lifeExp-line-point}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line() + geom_point()
```

It's important to note that each layer is drawn on top of the previous layer. In
this example, the points have been drawn *on top of* the lines. Here's a
demonstration:

```{r lifeExp-layer-example-1}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
  geom_line(aes(color=continent)) + geom_point()
```

In this example, the *aesthetic* mapping of **color** has been moved from the
global plot options in `ggplot` to the `geom_line` layer so it no longer applies
to the points. Now we can clearly see that the points are drawn on top of the
lines.

> ## Challenge 3 {.challenge}
>
> Switch the order of the point and line layers from the previous example. What
> happened?
>

There are many other geoms we can use to explore the data. One common one is a 
histogram, so we can see the distribution of a single variable:

```{r gdp-hist}
ggplot(gapminder, aes(x = gdpPercap)) + geom_histogram()
```

Or a boxplot to compare distribution of life expectancy across continents:
```{r life-exp-boxplot}
ggplot(gapminder, aes(x = continent, y = lifeExp)) + geom_boxplot()
```


## Transformations and statistics

ggplot also makes it easy to overlay statistical models over the data. To
demonstrate we'll go back to our first example:

```{r lifeExp-vs-gdpPercap-scatter3, message=FALSE}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color=continent)) +
  geom_point()
```

Currently it's hard to see the relationship between the points due to some strong
outliers in GDP per capita. We can change the scale of units on the y axis using
the *scale* functions. These control the mapping between the data values and
visual values of an aesthetic.

```{r axis-scale}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10()
```

The `log10` function applied a transformation to the values of the gdpPercap
column before rendering them on the plot, so that each multiple of 10 now only
corresponds to an increase in 1 on the transformed scale, e.g. a GDP per capita
of 1,000 is now 3 on the y axis, a value of 10,000 corresponds to 4 on the y
axis and so on. This makes it easier to visualise the spread of data on the
y-axis.

We can fit a simple linear relationship to the data by adding another layer,
`geom_smooth`:

```{r lm-fit}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm")
```

We can make the line thicker by *setting* the **size** aesthetic in the
`geom_smooth` layer:

```{r lm-fit2}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm", size=1.5)
```

There are two ways an *aesthetic* can be specified. Here we *set* the **size**
aesthetic by passing it as an argument to `geom_smooth`. Previously in the
lesson we've used the `aes` function to define a *mapping* between data
variables and their visual representation.

> ## Challenge 4 {.challenge}
>
> Modify the color and size of the points on the point layer in the previous
> example.
>
> Hint: do not use the `aes` function.
>

## Multi-panel figures

Earlier we visualised the change in life expectancy over time across all
countries in one plot. Alternatively, we can split this out over multiple panels
by adding a layer of **facet** panels:

```{r facet}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country)
```

The `facet_wrap` layer took a "formula" as its argument, denoted by the tilde
(~). This tells R to draw a panel for each unique value in the country column
of the gapminder dataset.

## Modifying text

To clean this figure up for a publication we need to change some of the text
elements. The x-axis is way too cluttered, and the y axis should read
"Life expectancy", rather than the column name in the data frame.

We can do this by adding a couple of different layers. The **theme** layer
controls the axis text, and overall text size, and there are special layers
for changing the axis labels. To change the legend title, we need to use the
**scales** layer.

```{r theme}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  xlab("Year") + ylab("Life expectancy") + ggtitle("Figure 1") +
  scale_fill_discrete(name="Continent") +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
```


This is just a taste of what you can do with `ggplot2`. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc].
Finally, if you have no idea how to change something, a quick google search will
usually send you to a relevant question and answer on Stack Overflow with reusable
code to modify!

[cheat]: http://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
[ggplot-doc]: http://docs.ggplot2.org/current/


> ## Challenge 5 {.challenge}
>
> Create a density plot of GDP per capita, filled by continent.
>
> Advanced:
>  - Transform the x axis to better visualise the data spread.
>  - Add a facet layer to panel the density plots by year.
>


### Further **ggplot2** resources

* [The official **ggplot2** documentation](http://docs.ggplot2.org/current/)
* [The **ggplot2** book, by the developer, Hadley Wickham](http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20)
* [The **ggplot2** Google Group (mailing list, discussion forum)](https://groups.google.com/forum/#!forum/ggplot2).
* [Intermediate Software Carpentry lesson on data visualization with **ggplot2**](https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization).
* [A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**](http://learnr.wordpress.com/).
* [Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site](http://stackoverflow.com/questions/tagged/ggplot2).

## Challenge solutions

Solutions to challenges

> ## Solution to challenge 1 {.challenge}
>
> Modify the example so that the figure visualise how life expectancy has
> changed over time:
>
> ```{r ch1-sol}
> ggplot(data = gapminder, aes(x = year, y = lifeExp)) + geom_point()
> ```
>

> ## Solution to challenge 2 {.challenge}
>
> In the previous examples and challenge we've used the `aes` function to tell
> the scatterplot **geom** about the **x** and **y** locations of each point.
> Another *aesthetic* property we can modify is the point *color*. Modify the
> code from the previous challenge to **color** the points by the "continent"
> column. What trends do you see in the data? Are they what you expected?
>
> ```{r ch2-sol}
> ggplot(data = gapminder, aes(x = year, y = lifeExp, color=continent)) +
>   geom_point()
> ```
>

> ## Solution to challenge 3 {.challenge}
>
> Switch the order of the point and line layers from the previous example. What
> happened?
>
> ```{r ch3-sol}
> ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
>  geom_point() + geom_line(aes(color=continent))
> ```
> 
> The lines now get drawn over the points!
>


> ## Solution to challenge 4 {.challenge}
>
> Modify the color and size of the points on the point layer in the previous
> example.
>
> Hint: do not use the `aes` function.
>
> ```{r ch4-sol}
> ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
>  geom_point(size=3, color="orange") + scale_y_log10() +
>  geom_smooth(method="lm", size=1.5)
> ```
>

> ## Solution to challenge 5 {.challenge}
>
> Create a density plot of GDP per capita, filled by continent.
>
> Advanced:
>  - Transform the x axis to better visualise the data spread.
>  - Add a facet layer to panel the density plots by year.
>
> ```{r ch5-sol}
> ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
>  geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
> ```
>
