# Data visualization with ggplot2

## Basic plotting

R's base (built-in) plotting functions are powerful and very flexible, but not 
overly user friendly. For simple exploratory plots that don't need to look nice, 
they are useful. They are generally specified  as `plot(x, y, ...)`

```{r baseplot1}
plot(dat$Sepal.Width, dat$Sepal.Length, type="p", col=dat$Species)
```

You can also specify them in a formula format `plot(y ~ x, data='', ...)`
```{r baseplot2}
plot(Sepal.Length ~ Sepal.Width, data=dat, col=Species)
boxplot(Sepal.Length ~ Species, data=dat)
```

For more advanced (and easier) plotting, we're going to use another package by 
Hadley Wickham, the same person who developed **dplyr**

## ggplot2

### The Diamonds dataset

Now let's look at a bigger dataset. We're going to be using a data visualization 
package called **ggplot2** for drawing the plots, and the **ggplot2** package 
comes with some data we're going to use for this example.

Recall how to install and load packages. Install the package if you haven't already:

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```

Now let's load the diamonds dataset and take a look at the first few rows and it's structure with commands we learned previously. To learn more about this dataset you can also run `?diamonds`. 

```{r diamondshead}
data(diamonds)
head(diamonds)
str(diamonds)
```

From this we can see this dataset has prices of nearly 54,000 diamonds along 
with various features about the diamonds, such as the weight, the quality of 
the cut, the color, the clarity, and measurements of various dimensions. 

If we wanted to do some exploratory data analysis we might start by plotting the 
price versus the weight of the diamond using base graphics.

```{r basediamond}
plot(price ~ carat, data=diamonds)
```

As we would expect there is definitely a relationship between the size of the 
diamond and its cost, but how do the other variables (cut, color, clarity) 
affect the price? We could examine the interrelationships of all these variables 
using base R graphics, but it could become extremely cumbersome. 

**ggplot2** is a widely used R package that extends R's visualization 
capabilities. It takes the hassle out of things like creating legends, mapping 
other variables to scales like color, or faceting plots into small multiples. 
We'll learn about what all these things mean shortly. To start with, let's 
produce the same plot as before, but this time using **ggplot2**.

### ggplot2 and the *Grammar of Graphics*

The **ggplot2** package provides an R implementation of Leland Wilkinson's 
*Grammar of Graphics* (1999). The *Grammar of Graphics* challenges data analysts 
to think beyond the garden variety plot types (e.g. scatter-plot, barplot) and 
to consider the components that make up a plot or graphic, such as how data are 
represented on the plot (as lines, points, etc.), how variables are mapped to 
coordinates or plotting shape or colour, what transformation or statistical 
summary is required, and so on. Specifically, **ggplot2** allows users to build 
a plot layer-by-layer by specifying:

* The _data_, plus some _aesthetics_ that map variables in the data to axes on 
  the plot or to size, shape, color, etc.,
* a _geom_, which specifies how the data are represented on the plot (points, 
  lines, bars, etc.),
* a _stat_, a statistical transformation or summary of the data applied prior to 
  plotting,
* _facets_, which we've already seen above, that allow the data to be divided 
  into chunks on the basis of other categorical or continuous variables and the 
  same plot drawn for each chunk.

Because **ggplot2** implements a *layered* grammar of graphics, data points and 
additional information (scatterplot smoothers, confidence bands, etc.) can be 
added to the plot via additional layers, each of which utilize further geoms, 
aesthetics, and stats.

To make the best use of **ggplot2** it helps to understand the grammar and how 
it affects how plots are produced. In addition, it is important to note that 
**ggplot2** is not a general-purpose plotting tool-kit; you may not be able to 
achieve certain plots or additions to a figure of they do not map onto concepts 
included in the layered grammar.

The `ggplot` function has two required arguments: the *data* used for creating 
the plot, and an *aesthetic* mapping to describe how variables in said data are 
mapped to things we can see on the plot. Let's use `ggplot` to recreate some of 
the same plots we produced above. First, the simple scatterplot:

```{r ggdiamonds}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point()
```

Here, we've built our plot in layers. First, we create a canvas for plotting 
layers to come using the `ggplot` function, specifying which **data** to use 
(here, the *diamonds* data frame), and an **aesthetic mapping** of *carat* to 
the x-axis and *price* to the y-axis. We next add a layer to the plot, 
specifying a **geom**, or a way of visually representing the aesthetic mapping. 
Here we're using a point, and the funtion is `geom_point()`. Parts are layered 
together using the `+` operator

Instead of using a point, we could use a different geom. Here, let's use 
[hexagonal binning](http://cran.r-project.org/web/packages/hexbin/vignettes/hexagon_binning.pdf) instead of a point.

```{r gghexbin}
ggplot(diamonds, aes(carat, price)) + geom_hex()
```

Here, each bin represents a segment of the plotting surface with lighter blue 
colors representing more density in that segment. The number of bins can be 
adjusted as an argument to the `geom_hex()` function. This is one method of 
solving the overplotting problem we have in this plot.

```{r gghexbin50}
ggplot(diamonds, aes(carat, price)) + geom_hex(bins=50)
```

Another method, here using points again, is to lower the opacity of each point. 
Here, `alpha=1/5` sets the opacity of each point to 20%. In other words, 5 
points would have to overlap to result in a completely solid point. Note that 
in this case we're not *mapping* the alpha level aesthetic to some other 
variable as we did above with color -- we're setting it to a static value of 
0.20 for all points in the layer.

```{r gghexbin2}
ggplot(diamonds, aes(carat, price)) + geom_point(alpha=1/5)
```

We can easily add more layers to the plot. For instance, we could add another 
layer displaying a smoothed conditional mean using the `geom_smooth()` function.

```{r smoothgam}
ggplot(diamonds, aes(carat, price)) + geom_point() + geom_smooth()
```

We'll get a message telling us that because we have >1,000 observations we will 
default to using a generalized additive model. We could easily plot a straight 
line by specifying that we want a linear model (`method="lm"`) instead of a 
generalized additive model, the default for large datasets.

```{r smoothlinear}
ggplot(diamonds, aes(carat, price)) + geom_point() + geom_smooth(method="lm")
```

We see again the strong relationship between the size of the diamond and its 
price, but the relationship doesn't appear linear. How does the diamond's 
clarity affect the weight-price relationship?

#### Faceting and scaling

One option we could use is to color-code the points by their clarity. Here, we 
add another *aesthetic*, colour (`col`) and map it to the variable we want it 
to represent: clarity

```{r ggcolclarity}
ggplot(diamonds, aes(x=carat, y=price, col=clarity)) + geom_point()
```

Examining the plot you can see that poor-clarity diamonds (included, small 
inclusions, etc) fetch a lower price per carat relative to more clear diamonds 
(very small inclusions, internally flawless, etc). We can see that **ggplot2** 
color-codes the points using a sensible default color scheme, and automatically 
draws a legend on the side for us. This requires a good deal of extra 
error-prone coding using base graphics.

However, with 54,000 points on this plot, there is a good deal of overplotting 
that obscures how clarity affects the nature of the weight-price relationship. 
How else might we visualize this data? This is where a *series of small multiples* 
is helpful. The idea of *small multiples* was popularized by data visualization 
expert Edward Tufte. The idea is that you create a large grid of small plots, 
where each plot shows a particular *facet* of the data. Here, each plot in the 
grid might be price vs. carat for each particular clarity level. You explain to 
your audience the axes and how to interpret each plot only once, and the 
audience will immediately understand the rest of the plots.

This can be accomplished easily using **ggplot2**:

```{r facetclar}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point() + facet_wrap( ~ clarity)
```

Here, the `facets` argument expects a forumla object that's constructed with 
the `~` operator. Here, we've plotted the price vs. weight separately for each 
level of clarity. We can see what we suspected before. With dirty diamonds 
(included, and perhaps small inclusions), the weight-price relationship is 
linear or slightly quadratic. Large diamonds can be purchased rather cheaply. 
But for very clear diamonds (internally flawless), the relationship is quadratic 
or even exponential.

Let's examine the weight-price relationship for various color ratings:

```{r facetcol}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point() + facet_wrap( ~ color)
```

Here we see that for whiter diamonds (D, E, F) the price rises more quickly with 
increasing weight than for yellower diamonds (H, I, J).

We can further facet the plot across two different categorical variables using 
the same syntax:

```{r ggfacet, fig.height=10, fig.width=10}
ggplot(diamonds, aes(carat, price)) + geom_point() + facet_grid(clarity ~ color)
```

Here we see that the price per carat rises very steeply for very white, very 
clear diamonds, while the relationship is nearly linear for yellower, more 
flawed diamonds. We can see that a perfect white diamond averages around $15,000 
while a yellow included diamond can be had for only around $2,000.

Finally, we can combind both color-coding and faceting in the same plot. Let's 
use the same faceting scheme as last time, but color the points by the quality 
of the diamond's cut.

```{r ggfacetclarcol_colcut, fig.height=10, fig.width=10}
ggplot(diamonds, aes(x=carat, y=price, col=cut)) + 
  geom_point() + 
  facet_grid(clarity ~ color)
```

This color-coding reveals that clearer, whiter diamonds *generally* have higher 
quality cuts, but the relationship doesn't appear strong, visually. Looking down 
the plot toward clearer diamonds you start to see more "Ideal" cuts than at the 
top, which are the more included diamonds.

What we've done here in addition to faceting is map a feature of the data (here, 
the cut quality) onto a scale (here, color). This behavior will work differently 
depending on whether you're looking at categorical or continuous variables. We 
can also map features to other *scales* such as `size=`, `shape=`, `linetype=`, 
or even transparency using `alpha=`. All of these different scales can be 
combined with each other or with facets, and give you an extremely powerful and 
easy-to-use graphical toolbox for exploratory data analysis.

By combining multiple layers with aesthetic mappings to different scales, 
**ggplot2** provides a foundation for producing a wide range of statistical 
graphics beyond simple "named" plots like scatter plots, histograms, bar plots, 
etc.

How about a stacked histogram, mapping the fill color of the stacked histogram 
to levels of the *clarity* variable:

```{r ggfillhisto}
ggplot(diamonds, aes(price, fill=clarity)) + 
  geom_histogram(position="fill", binwidth=200)
```

Or what about box plots of the price grouped separately by the quality of the 
cut, color-coded by the color of the diamond, with the price on the y-axis 
being on the log (base 10) scale? Simple:

```{r ggboxplots, fig.width=12}
ggplot(diamonds, aes(cut, price)) + geom_boxplot(aes(fill=color)) + scale_y_log10()
```

Or what about a kernel density plot (think about a smooth histogram) of the 
diamond's depth in different semitransparent curves with the color fill 
mapped to each level of *cut*, limited to depths between 55 and 70, with a 
title and a proper axis labels? This also shows the syntax of 
building up a plot one step at a time. We first initialize the plot with 
`ggplot`, giving it the data we're working with, and aesthetic mappings. We 
then add a `geom_density` layer, limit the x-axis displayed, and finally give 
it a title and axis labels. The plot is in the **g** object here; we can simply 
enter `g` and the plot will be displayed.

```{r ggdepthdensity, warning=FALSE}
g <- ggplot(diamonds, aes(depth, fill=cut)) 
g <- g + geom_density(alpha=1/4) 
g <- g + xlim(55, 70)
g <- g + ggtitle("Depths by Cut Quality")
g <- g + xlab("Depth") + ylab("Density")
g
```

Finally, we can save the plot created using the `ggsave` function:

```{r ggsave, eval=FALSE}
ggsave(filename="table-depth-density.png", plot=g)
```

There are endless ways to combine aesthetic mappings with different geoms and 
multiple layers. Read about other **geom**s, mappings, scales, and other layer 
options at the links below.

#### Exercise

Now, plot a graph using the iris dataset using **ggplot2**. You don't have to 
use points; other geoms include: `gemo_bar`, `geom_area`, `geom_boxplot`, 
and many more.  Try to incorporate at least one extra aesthetic (over `x` and 
`y`). Some examples of aesthetics are `col`, `fill`,  `size`, `shape`, `alpha`, 
`linetype`.

```{r irisexamples}
ggplot(dat, aes(x=Petal.Length, y=Petal.Width, col=Sepal.Width, size=Sepal.Length)) + 
  geom_point() + 
  geom_smooth(method="lm") + 
  facet_wrap(~ Species)

ggplot(dat, aes(x=Petal.Length)) + geom_histogram() + facet_wrap(~Species)

ggplot(dat, aes(x=Petal.Length, fill=Species)) + geom_histogram()

ggplot(dat, aes(x=Species, y=Sepal.Length)) + geom_boxplot()

ggplot(dat, aes(x=Species, y=Sepal.Width, fill=Species)) + geom_violin()

```

What if we want to split up the measurements and compare them all?  We need the 
type of measurement (`Petal.length`, `Petal.Width`) as a variable rather than each 
as a column heading. There is another package by Hadley Wickham (see a trend 
here?) that helps us restructure, or reshape our data: It's called `reshape2`.

```{r irisreshapebox, warning=FALSE}
require(reshape2)
dat_long <- melt(dat)
head(dat_long)

ggplot(dat_long, aes(x=variable, y=value, fill=Species)) + geom_boxplot()

ggplot(dat_long, aes(x=variable, y=value, fill=Species)) + geom_boxplot()

ggplot(dat_long, aes(x=Species, y=value, fill=Species)) + 
  geom_violin() + 
  facet_wrap(~ variable) + 
  ylab("Measurement (cm)")
```

## Further **ggplot2** resources

* [The official **ggplot2** documentation](http://docs.ggplot2.org/current/)
* [The **ggplot2** book, by the developer, Hadley Wickham](http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20)
* [The **ggplot2** Google Group (mailing list, discussion forum)](https://groups.google.com/forum/#!forum/ggplot2).
* [Intermediate Software Carpentry lesson on data visualization with **ggplot2**](https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization).
* [A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**](http://learnr.wordpress.com/).
* [Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site](http://stackoverflow.com/questions/tagged/ggplot2).
