---
title: Functions
---

```{r, echo=FALSE}
require(knitr)
opts_chunk$set(results="markup")
```

## Objectives

* Define a function that takes arguments (parameters).
* Return a value from a function.
* Set default values for function parameters.
* Explain why we should divide programs into small, single-purpose functions.
* Defining a function

Let's start by defining a function `fahr_to_kelvin()` that converts temperatures from Fahrenheit to Kelvin:

```{r}
fahr_to_kelvin <- function(temp) {
    kelvin <- ((temp - 32) * (5 / 9)) + 273.15
    kelvin
}
```

The definition opens with the name of your new function, which is followed by the call to make it a `function()` and a parenthesized list of arguments. You can have as many input parameters as you would like (but too many might be bad style). The body, or implementation, is surrounded by curly braces `{ }`. In many languages, the body of the function - the statements that are executed when it runs - must be indented, typically using 4 spaces. While this is not a mandatory requirement in R coding, we strongly recommend you to adopt this as good practice.

When we call the function, the values we pass to it are assigned to those variables so that we can use them inside the function. The last line within the function is what R will evaluate as a returning value. Remember that the last line has to be a command that will print to the screen, and not an object definition, otherwise the function will return nothing - it will work, but will provide no output. For example, let's try running our function. Calling our own function is no different from calling any other function:

```{r}
fahr_to_kelvin(32)
paste('boiling point of water:', fahr_to_kelvin(212))
```

We've successfully called the function that we defined, and we have access to the value that we returned. However, it the function was redefined as follows

```{r}
fahr_to_kelvin <- function(temp) {
    kelvin <- ((temp - 32) * (5/9)) + 273.15
}
```

Now typing

```{r}
fahr_to_kelvin(32)
```

Will return nothing. Why?



## Composing Functions

Now that we've seen how to turn Fahrenheit into Kelvin, it's easy to turn Kelvin into Celsius:

```{r}
kelvin_to_celsius <- function(temp) {
    celsius <- temp - 273.15
        celsius
}

paste('absolute zero in Celsius:', kelvin_to_celsius(0))
```

What about converting Fahrenheit to Celsius? We could write out the formula, but we don't need to. Instead, we can compose the two functions we have already created:

```{r}
fahr_to_celsius <- function(temp) {
    temp_k <- fahr_to_kelvin(temp)
    result <- kelvin_to_celsius(temp_k)
    result
}

paste('freezing point of water in Celsius:', fahr_to_celsius(32.0))
```

This is our first taste of how larger programs are built: we define basic operations, then combine them in ever-large chunks to get the effect we want. 
Real-life functions will usually be larger than the ones shown here - typically half a dozen to a few dozen lines - but they shouldn't ever be much longer than that, or the next person who reads it won't be able to understand what's going on. __Modular programming__


### Challenges

As we've seen in our print statements, we can use `paste` to concatenate strings, `paste(a, b, sep = "")` is `ab`. __Note__: the `sep` can be an important value to define! What is the default? What can `sep` be?

1. Write a function called `fence` that takes two arguments called `original` and `wrapper` and returns a new string that has the `wrapper` character at the beginning and end of the `original`:

```{r, echo=FALSE, eval=TRUE}
fence  <-  function(original, wrapper) {
    paste(wrapper, original, wrapper, sep="")
}
```

```{r, results="markup"}
fence("name", "*")
```

## String splits

If the variable s refers to a string, then we can parse the string into its separate components - each of the characters. Base R has a function called `strsplit` that can be used to break up strings, into smaller chunks. 

```{r}
pangram <- "the quick brown fox jumps over the lazy dog"
strsplit(pangram, " ")
```
                
The output from strsplit is in a list. 
Notice that the unusual first line of `strsplit()`’s output consists of `[[1]]`. 
Similar to the way that R displays vectors, `[[1]]` means that R is showing the first element of a list. 
Lists are extremely important concepts in R; they allow you to combine all kinds of variables.
For example, a list can be made up of many elements, and elements could be vectors, data frames, matrices, or further lists. 

In this example, this list has only a single element. Yes, that’s right: The list has one element, but that element is a vector.

To extract an element from a list, you have to use double square brackets: `[[`. 
Split your pangram into words, and assign the first element to a new variable called words, using double-square-brackets `[[]]` subsetting, as follows:

```{r}
words <- strsplit(pangram, " ")[[1]]
```

We can then pull out the different words using indexing, where `words[1]` is the first element in the vector of `words` and `words[9]` would be the last:

```{r}
words[2]
words[9]
```

1. Write a function `out()` that returns a string made up of just the first and last characters of its input.
    a. Outline the steps you need to take to write this function. Discuss with the person sitting next you.
    b. Write part of the code, make sure it works.
    c. Write the next step.
    d. Test your function.
    e. Can your function handle words of different lengths?

```{r}
out <- function(word) {
    letter  <- strsplit(word, "")[[1]]
    abbrev  <- paste(letter[1], letter[length(letter)], sep="")
    abbrev
}

out('helium')
```

__Making our function work with different inputs__. If we want just the last word, but we can't remember how long our sentence is, we can use `length()`

```{r}
length(words)
words[length(words)]
```

__BREAK__

## Explaining the R Environments

Let's take a closer look at what happens when we call `fahr_to_celsius(32)`. To make things clearer, we'll start by putting the initial value 32 in a variable and store the final result in one as well:

```{r}
original <- 32
final    <- fahr_to_celsius(original)
```

_Discuss and draw a diagram showing what memory looks like after the first line has been executed. Point to the environment_

When we call `fahr_to_celsius()`, R creates a new environment called the *evaluation environment* that is *local* to the function. This environment consists of two things

 1. a *frame*, which contains the symbol-value pairs (the assignment, called *binding*, of a value to a variable),
 2. an *enclosure*, which points to the enclosing environment, i.e. the environment where the function was called.

Initially, the *frame* only holds the object `temp`. Until is is used, `temp` in *unevaluated*; in effect it is a placeholder for whatever value we passed to the `temp` argument, which gets resolved as needed. This is known as *lazy evaluation*.

*Scoping* refers to the rules by which languages look up the value of a variable (symbol). R has two type of scoping rules

 1. *lexically* scoping, and
 2. *dynamic* scoping.

We won't discuss *dynamic* scoping heres. Basically, *lexical* scoping means that R looks up values for variables (symbols) based on how functions were nested when they were *called*. If a name isn't defined inside a function, R will look for the name in in the *parent frame*, the frame in the calling environment that is one where the function was called

```{r, lexical-scoping}
a <- 10
f <- function() {
  b <- 5
  a * b
}
f()
```

There is no name `a` defined within the body of `f()`. Following lexical scoping rules, R will look in the *parent frame* of `f()` for a name `a`, which is found and has value `10`. As `f()` is defined in the workspace, the enclosing environment of `f()` is the *global environment*. Regardless of how functions are nested when called, R will always look for a name from the inside out until the global environment is reached.

When we nest a call to `fahr_to_kelvin()` inside `fahr_to_celsius()`, R creates another local environment to hold the variables involved in the evaluation of `fahr_to_kelvin()`.

When `fahr_to_celsius()` is called, it in turn calls `fahr_to_kelvin()`. As `temp` is passed to `fahr_to_kelvin()`, R evaluates `temp` to derive its value and this variable is passed to `fahr_to_kelvin()`. Now there are two `temp`s in play

 1. the `temp` local to `fahr_to_kelvin()`, and
 2. the `temp` local to `fahr_to_celcius()`.

As far as `fahr_to_kelvin()` is concerned there is only one `temp`; the `temp` local to it *masks* the other `temp`. That R knows which versions of variables with the same name belong to which function is due to environments.

When the call to `fahr_to_kelvin()` returns a value R throws away `fahr_to_kelvin()`'s frame and creates a new variable `temp_k` in the frame for `fahr_to_celsius()` to hold the temperature in Kelvin.

`kelvin_to_celsius()` is then called and a new environment to hold that function's variables is created.

Once again, R throws away that stack frame when `kelvin_to_celsius()` is done and creates the variable `result` in the frame of `fahr_to_celsius()`

Finally, when `fahr_to_celsius()` returns, R throws away its environment and puts `result` in a new variable called `final` that lives in the *global environment* where we called `fahr_to_celcius()` in the first place.

The summary of this is that the parent frame (the global environment is the parent frame of `fahr_to_celsius()`, the local environment of `fahr_to_celsius()` is the parent frame of `fahr_to_kelvin()`) is the environment where a function was defined (lexical scoping), the parent frame is the frame/environment from which the function was called.

The *global environment* is the final environment/frame and is always present; it holds the variables we defined outside the functions in our code. What it doesn't hold are the variables created during function calls. If we try to get the value of `temp` *after* our functions have finished running, R tells us that there's no such thing:

```{r eval=FALSE}
paste('final value of temp after all function calls:', temp)
```

Why go to all this trouble? Well, here's a function called `span()` that calculates the difference between the mininum and maximum values in an array:

```{r}
span <- function(a) {
    diff <- max(a) - min(a)
    diff
}
```

Notice `span()` assigns a value to variable called `diff`. We might very well use a variable with the same name (`diff`) to hold data:

```{r}
diff <- c(46, 55, 26, 64, 31, 68, 100, 79, 39, 95)
span(diff)
```

We don't expect the variable `diff` to have the value `r span(diff)` after this function call, so the name `diff` cannot refer to the same variable defined inside `span()` as it does in your workspace (the global environment). And yes, we could probably choose a different name than `diff` for our variable in this case, but we don't want to have to read every line of R code we write that we use to see what variable names its functions use before calling any of those functions, just in case they change the values of our variables.

The big idea here is __encapsulation__, and it's the key to writing correct, comprehensible programs. A function's job is to turn several operations into one so that we can think about a single function call instead of a dozen or a hundred statements each time we want to do something. That only works if functions don't interfere with each other; if they do, we have to pay attention to the details once again, which quickly overloads our short-term memory.
