---
title: Getting R Help
---

To get help for a particular function in R, type `?` and then the function name:

```{r eval=FALSE}
?mean
```

To get help for a topic in R, do a "fuzzy"" search with `??` (wrap the phrase in 
quotes if more than one word):

```{r eval=FALSE}
??"t-test"
```

### General help

* [Cookbook for R](http://www.cookbook-r.com/) - lots of plotting help here, 
  including ggplot2
* Google (It actually knows what "R" is now)
* [Stackoverflow](http://stackoverflow.com/questions/tagged/r) (use the `[r]` tag - also `[ggplot2]`, `[dplyr]`, etc.)
* http://www.rdocumentation.org/
* Each other! (Talk it out)
* [Tell it to the duck](http://blog.codinghorror.com/rubber-duck-problem-solving/)

### ggplot2

* [ggplot2 official documentation](http://docs.ggplot2.org/current/)
* [Color Brewer](http://colorbrewer2.org/): A great general resource for choosing colour palettes

### Books

* [The Art of R Programming]()
* [Hadley Wickham's online book: Advanced R Programming](http://adv-r.had.co.nz/)
* [The R Graphics Cookbook](http://www.amazon.com/gp/product/1449316956) by 
  Winston Change - The paper version of the Cookbook for R website mentioned above
  
### Other learning resources - online courses, etc

* [R for cats](http://rforcats.net/)
* [Try R - codeschool](http://tryr.codeschool.com/)
* [swirl](http://swirlstats.com/) - Learn statistics and R simultaneously - within R itself!
* [R Programming (Coursera)](https://www.coursera.org/course/rprog)
    * There are a number of R and/or statistics courses on Coursera.  It's a free 
      online learning platform, and the courses are taught by high calibre 
      professors from good universities.
