---
---

# Repeating Things

** The structure and much of the content in this module was borrowed from 
[Software Carpentry](http://www.softwarecarpentry.org)'s
[novice R Bootcamp material](https://github.com/swcarpentry/bc/tree/master/novice/r) 
(Copyright (c) Software Carpentry), which they make available for reuse under the 
Creative Commons Attribution [(CC_BY)](http://creativecommons.org/licenses/by/3.0/) 
license.

One of the big advantages of working with data in an environment like R is that 
we can repeat and automate things really easily.

## *apply family of functions: `sapply()` and `lapply()`

These functions _apply_ an arbitrary function over a list (or vector).  They can 
be a little tough to master, but they can be really useful.

### Lists

In R lists are a lot like vectors. Unlike vectors however, the contents of a 
list are not restricted to a single data type and can encompass any mixture of 
data types (even other lists!). This makes them fundamentally different from 
vectors.

Create lists using `list()` or coerce other objects using `as.list()`

```{r}
x <- list(1, "a", TRUE, 1+4i)
x

x <- 1:10
x <- as.list(x)
length(x)
```


Lists, like vectors, can be _indexed_, though slightly differently.  Use double 
square brackets `list[[index]]` to get the contents of a list element. Using 
single square will still return a list.

1. What is the class of `x[1]`?
2. How about `x[[1]]`?

```{r}
Andy <- list(name = "Andy", fav_nums = 1:10, fav_data = head(iris))
Andy
```

1. What is the length of this object? What about its structure?

Lists can be extremely useful inside functions. You can "staple" together lots 
of different kinds of results into a single object that a function can return.

A list does not print to the console like a vector. Instead, each element of 
the list starts on a new line.

* A data frame is a special type of list where every element of the list is a 
vector of the same length.

### lapply

What it does: Returns a list of same length as the input. 
Each element of the output is a result of applying a function to the 
corresponding element.

```{r, eval = TRUE}
my_list <- list(a = 1:10, b = 2:20)
my_list
lapply(my_list, mean)
```

### sapply

`sapply` is a more user friendly version of `lapply` and will simplify its 
output to a vector if it can.

Let's work with the same list we just created.

```{r, eval = TRUE}
my_list
sapply(my_list, mean)
```

Now lets' see where this might actually be useful.

Say instead of all of our iris data in one nice file, we have it in separate 
files; we could read them in one by one, but what if we had 50 files?  It would 
get really tedious.  First, we can list the files with the `list.files()` 
function:

```{r}
list.files("data")
list.files("data", pattern = "iris_", full.names=TRUE)
iris_files <- list.files("data", pattern = "iris_", full.names=TRUE)
```

We can now take that list of files, and using `lapply()` and the `read.csv()` 
function to read in all of those files.

```{r}
iris_list <- lapply(iris_files, read.csv, stringsAsFactors=FALSE)
dat2 <- rbind_all(iris_list)
```

## For loops

```{r}
x <- c("apples", "oranges", "bananas", "strawberries")

for(i in x) {
    print(i)
}

for(i in 1:4) {
    print(x[i])
}

for(i in seq_along(x)) {
    print(x[i])
}
```

Let's say we had our iris data in one big file (which we do), but we needed to 
save a file for each species.

```{r eval=FALSE}
for (name in unique(dat$Species)) {
  subdat <- dat[dat$Species == name,]
  write.csv(subdat, paste("data/iris_", name, ".csv", sep=""), 
            row.names = FALSE)
}
```

