---
---

# Data munging with dplyr

** Most of the content in this module came from the dplyr 
[Introduction vignette](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) 
(Copyright (c) 2013 Hadley Wickham, RStudio).

While a lot of powerful tools are built into R, its greatest power comes from 
its massive base of users and developers, and the tools they create.  These 
tools come in the form of packages that we download, then load - they give us 
access to many more functions.  Loading a package into R is like getting a piece 
of lab equipment out of a storage locker and setting it up on the bench. Once 
it's done, we can ask the package to do things for us.

We're going to start by using a package called "dplyr", built by Hadley Wickham.  
dplyr provides a set of simple functions that correspond to the most common data 
manipulation verbs, so that you can easily translate your thoughts into code. 
Most of the this functionality is available in base **R**, but this package 
makes this kind of work much easier.

```{r eval=FALSE}
install.packages("dplyr") # Install the package
```

```{r}
require(dplyr) # Load the package into memory
```

`dplyr` implements the following verbs useful for data manipulation:

* `group_by()`: set the grouping variable(s)
* `filter()`: focus on a subset of rows
* `select()`: focus on a subset of variables
* `mutate()`: add new columns (also `mutate_each()`)
* `summarise()`: reduce each group to a smaller number of summary statistics 
  (also `summarise_each()`)
* `arrange()`: re-order the rows

## Filter rows with `filter()`

`filter()` allows you to select a subset of the rows of a data frame. The first 
argument is the name of the data frame, and the second and subsequent are 
filtering expressions evaluated in the context of that data frame

```{r}
filter(dat, Sepal.Length > 5, Petal.Length > 1.5)
```

Remember we talked about indexing to select rows... the equivalent would be:

```{r}
dat[dat$Sepal.Length > 5 & dat$Petal.Length > 1.5,]
```

What if we want to select flowers with Sepal Length > 5 **OR** Petal Length > 1.5?
(The above uses **AND**).  In **R**, the `|` (pipe) character means **OR**

```{r}
filter(dat, Sepal.Length > 5 | Petal.Length > 1.5)
## Works with indexing as well:
dat[dat$Sepal.Length > 5 | dat$Petal.Length > 1.5,]
```

## Arrange rows with `arrange()`

`arrange()` works similarly to `filter()` except that instead of filtering or 
selecting rows, it reorders them. It takes a data frame, and a set of column 
names (or more complicated expressions) to order by. If you provide more than 
one column name, each additional column will be used to break ties in the values 
of preceding columns:

```{r}
arrange(dat, Sepal.Length, Sepal.Width)
```

Use `desc()` to order a column in descending order:

```{r}
arrange(dat, desc(Petal.Length))
```

## Select columns with `select()`

Often you work with large datasets with many columns where only a few are 
actually of interest to you. `select()` allows you to rapidly zoom in on a 
useful subset using operations that usually only work on numeric variable positions:

```{r}
# Select columns by name
select(dat, Sepal.Length, Sepal.Width, Species)
# Select all columns between Petal.Length and Species (inclusive)
select(dat, Petal.Length:Species)
# Select all columns except the last two (inclusive)
select(dat, -(Sepal.Length:Sepal.Width))
```

## Add new columns with `mutate()`

As well as selecting from the set of existing columns, it's often useful to add 
new columns that are functions of existing columns. This is the job of `mutate()`:

```{r}
mutate(dat,
  Petal.Ratio = Petal.Length / Petal.Width,
  Sepal.Ratio = Sepal.Length / Sepal.Width)
```

## Summarise values with `summarise()`

The last verb is `summarise()`, which collapses a data frame to a single row. 
It's not very useful (yet):

```{r}
summarise(dat, mean.Petal.Length = mean(Petal.Length, na.rm = TRUE))
```

_Notice that we haven't been assigning these things to any objects, so they just 
print out to the console and then are gone.

You may have noticed that all these functions are very similar:

* The first argument is a data frame.

* The subsequent arguments describe what to do with it, and you can refer
  to columns in the data frame directly without using `$`.

* The result is a new data frame

These five functions provide the basis of a language of data manipulation. At 
the most basic level, you can only alter a tidy data frame in five useful ways: 
you can reorder the rows (`arrange()`), pick observations and variables of 
interest (`filter()` and `select()`), add new variables that are functions of 
existing variables (`mutate()`) or collapse many values to a summary 
(`summarise()`). The remainder of the language comes from applying the five 
functions to different types of data, especially to grouped data using the 
`group_by()` function.

It's used for the **split, apply, combine** approach to data manipulation: You 
**split** a dataset by some variable, **apply** some operation(s) to each piece, 
and **combine** the pieces back together.

We'll use this to generate some _per-species_ statistics, because the mean of 
some measurement across all species probably isn't that helpful - like getting 
the average length of tails of lions and house cats.

```{r}
grp_spp <- group_by(dat, Species)
grp_spp

summarise(grp_spp, 
          mean.Petal.Length = mean(Petal.Length), 
          sd.Petal.Length = sd(Petal.Length))

mutate(grp_spp, SL.mean.ratio = Sepal.Length / mean(Sepal.Length))
```

This is nice and handy, but what if we want to do several operations that depend 
on the result of a previous operation? We want to group by species, use only the 
individuals that have a long Sepal Length (greater than the mean for that 
species), calculate the number of each species that meet those criteria, then 
calculate the mean and standard deviation of petal length for each species.

```{r}
grp_spp <- group_by(dat, Species)
temp1 <- filter(grp_spp, Sepal.Length > mean(Sepal.Length))
temp2 <- mutate(temp1, n=n()) # n(): number of observations in the current group
summary_dat <- summarise(temp2, 
                        mean.Petal.Length = mean(Petal.Length), 
                        sd.Petal.Length = sd(Petal.Length))

```

you can also chain these things together by using the `%>%` operator. This helps 
make your code very readable, and eliminates the need to create intermediate 
objects.

`x %>% f(y)` turns into `f(x, y)` so you can use it to write multiple 
operations so you can read from left-to-right, top-to-bottom:

```{r}
sum_dat <- dat %>%
  group_by(Species) %>%
  filter(Sepal.Length > mean(Sepal.Length)) %>%
  mutate(n = n()) %>%           # n(): number of observations in the current group
  summarise(mean.Petal.Length = mean(Petal.Length), 
            sd.Petal.Length = sd(Petal.Length))
```

### Exercise

Summarise `dat` by species, returning a data frame that calculates the median 
ratio of sepal length and sepal width.

```{r}
sep_ratio <- dat %>%
  group_by(Species) %>%
  mutate(sep_diff = Sepal.Length / Sepal.Width) %>%
  summarise(med_sep_diff = median(sep_diff, na.rm=TRUE))
```


Now let's clean up a bit. See all objects in your workspace with `ls()`.  You 
also see these things in your environment pane in RStudio.

```{r eval=FALSE}
ls()
```

Remove them with `rm`:

```{r}
rm(temp1)
# Or remove a list of objects:
rm(temp2, summary_dat)
```

## Outputting files

We've done our summaries, but now we want to save them so we can share our 
amazing findings.  Just like we use the function `read.csv()` to read in data, 
we use `write.csv()` to write it out.

```{r eval=FALSE}
write.csv(sum_dat, "iris_summaries.csv", row.names=FALSE)
```

