---
title: "An Introduction to R"
author: "Andy Teucher"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 1
---

# Introduction

### Course website
These notes are a pdf version of the website for the course, which can be viewed at:
https://ateucher.github.io/rcourse_site

### Credits
Most of the material here was borrowed and adapted from 
[Software Carpentry](http://www.softwarecarpentry.org)'s
[novice R Bootcamp material](https://github.com/swcarpentry/r-novice-gapminder), 
which they make available for reuse under the Creative Commons Attribution [(CC_BY)](http://creativecommons.org/licenses/by/3.0/) license. 
These are amazing people, doing amazing things to help the scientific world be 
more productive with their code and data. Check them out, and if you get the chance 
to attend a bootcamp, do it.

The [course notes](http://www.poissonconsulting.ca/courses/2012/06/05/an-introduction-to-r-course-12.html) from [Poisson Consulting](http://www.poissonconsulting.ca/)'s 
2012 R Course were also very helpful in putting this material together.

### Source
Source material for the course notes can be found here: https://github.com/ateucher/rcourse_site

### License
Notes: [CC-BY](http://creativecommons.org/licenses/by/3.0/). 
Code: [MIT](http://opensource.org/licenses/mit-license.html)

# Introduction to R and RStudio

```{r child = "../01-introduction.Rmd"}
```

# Data frames

```{r child = "../02-data_frames.Rmd"}
```

# Data visualization with ggplot2

```{r child = "../03-plotting.Rmd"}
```

# Subsetting data

```{r child = "../04-subsetting.Rmd"}
```

# Dataframe manipulation with dplyr

```{r child = "../05-dplyr.Rmd"}
```

# Dataframe manipulation with tidyr

```{r child = "../06-tidyr.Rmd"}
```

# Writing data

```{r child = "../07-writing-data.Rmd"}
```

# Basic statistics

```{r child = "../08-basic_stats.Rmd"}
```

# Writing functions

```{r child = "../09-functions.Rmd"}
```

# Flow control

```{r child = "../10-control-flow.Rmd"}
```

# Best Practices

```{r child = "../best-practices.Rmd"}
```

# Getting help

```{r child = "../getting_help.Rmd"}
```
